---
title: >
  Using unisets with Bioconductor annotation packages
author:
- name: Kevin Rue-Albrecht
  affiliation: 
  - &id4 Kennedy Institute of Rheumatology, University of Oxford,
    Headington, Oxford OX3 7FY, UK.
  email: kevin.rue-albrecht@kennedy.ox.ac.uk
- name: Robert A. Amezquita
  affiliation: 
  - &id2 Fred Hutchinson Cancer Research Center, Seattle, WA, USA, 98115
  email: robert.amezquita@fredhutch.org
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('unisets')`"
output: 
  BiocStyle::html_document:
    toc_float: true
abstract: |
  Importing gene set relationships from Bioconductor annotation packages.
vignette: |
  %\VignetteIndexEntry{2. Using unisets with Bioconductor annotation packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
  %\VignettePackage{unisets}
---

# Getting started

## Bioconductor annotation packages

Annotation packages are available from Bioconductor for a range of model species.
Users may browse [BiocViews "AnnotationData"](https://bioconductor.org/packages/release/BiocViews.html#___AnnotationData)
on the Bioconductor website or search packages programmatically using the command below.

```{r, eval=FALSE}
BiocManager::available("^org\\.")
```

Here, we load the human gene annotations.

```{r message=FALSE}
library(org.Hs.eg.db)
```

# Converting to unisets classes

`Go3AnnDbBimap` objects (from the `r Biocpkg("AnnotationDbi")` package) are maps between Entrez gene identifiers and Gene Ontology (GO) identifiers.
Those objects may be directly converted to `BaseSets` objects as demonstrated below.

```{r message=FALSE}
library(unisets)
base_sets <- import(org.Hs.egGO)
relations(base_sets)
```

Notice how the `"element"` information is typed as `EntrezIdVector`, allowing the type of identifier to affect downstream methods (e.g., pathway analyses).
The `EntrezIdVector` class directly inherits from the `IdVector` class, and benefits of all the methods associated with the parent class.

It is also useful to note that the conversion of `Go3AnnDbBimap` Gene Ontology maps to `r Githubpkg("kevinrue/unisets")` objects automatically fetches metadata for each GO identifier from the `GO.db` package, if installed.
The metadata is stored it in the `mcols` (metadata-columns) slot of the `setData` slot of the object returned.
This metadata can be accessed using the accessor method of the same name.

```{r}
mcols(setData(base_sets))[, c("ONTOLOGY", "TERM")]
```

We may then visualize the distribution of set sizes, on a log~10~ scale.

```{r message=FALSE}
library(ggplot2)
library(cowplot)
ggplot(data.frame(setLengths=setLengths(base_sets))) +
    geom_histogram(aes(setLengths), bins=100, color="black", fill="grey") +
    scale_x_log10() + labs(y="Sets", x="Genes")
```

`org.Hs.egGO` is an R object that provides mappings between entrez gene identifiers and the GO identifiers that they are directly associated with.
This mapping and its reverse mapping do NOT associate the child terms from the GO ontology with the gene.
Only the directly evidenced terms are represented here.

In contrast, `org.Hs.egGO2ALLEGS` is an R object that provides mappings between a given GO identifier and all of the Entrez Gene identifiers annotated at that GO term OR TO ONE OF IT'S CHILD NODES in the GO ontology.
Thus, this mapping is much larger and more inclusive than `org.Hs.egGO2EG`.

Below, we use the `length` method to show the number of relations between genes and GO terms imported from the `org.Hs.egGO2ALLEGS` map.

```{r}
base_sets <- import(org.Hs.egGO2ALLEGS)
format(length(base_sets), big.mark=",")
```

We can also examine the count of relations associated with each evidence code in each Gene Ontology namespace.

```{r, fig.height=7}
ggplot(as.data.frame(base_sets)) +
    geom_bar(aes(Evidence)) + facet_wrap(~Ontology, ncol = 1) + coord_flip() +
    # scale_y_continuous(labels = function(x){ format(as.integer(x), big.mark = ",") }) +
    scale_y_continuous(labels = scales::comma) +
    theme(axis.text.y = element_text(size=rel(0.7)))
```

# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
