---
title: >
  An introduction to the unisets package 
author:
- name: Kevin Rue-Albrecht
  affiliation: 
  - &id1 Kennedy Institute of Rheumatology, University of Oxford,
    Headington, Oxford OX3 7FY, UK.
  email: kevin.rue-albrecht@kennedy.ox.ac.uk
- name: Robert A. Amezquita
  affiliation: 
  - &id2 Fred Hutchinson Cancer Research Center, Seattle, WA, USA, 98115
  email: robert.amezquita@fredhutch.org
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('unisets')`"
output: 
  BiocStyle::html_document:
    toc_float: true
abstract: |
  Introduction to the unisets package.
vignette: |
  %\VignetteIndexEntry{1. An introduction to the unisets package}
  %\VignetteEncoding{UTF-8}  
  %\VignettePackage{unisets}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

# Overview

The goal of the `r Githubpkg("kevinrue/unisets")` package is to provide a collection of S4 classes to store relationships between elements and sets, with a particular emphasis on gene sets.

# Getting started

## The _BaseSets_ class

This is a basic example which shows you how to create a `BaseSets` object, to store simple associations between genes and sets, along with optional metadata associated with each relation:

```{r, message=FALSE}
library(unisets)
gene_lists <- list(
    geneset1 = c("A", "B"),
    geneset2 = c("B", "C", "D")
)
relations_table <- DataFrame(
    element = unlist(gene_lists),
    set     = rep(names(gene_lists), lengths(gene_lists)),
    extra1  = sample(c("ABC", "DEF"), 5L, replace=TRUE),
    extra2  = rbinom(5L, 10L, 0.4)
)
base_sets <- BaseSets(relations_table)
base_sets
```

Metadata for each element and set can be provided as separate `DataFrame` objects:

```{r, message=FALSE}
gene_data <- IdVector(c("A", "B", "C", "D"))
mcols(gene_data) <- DataFrame(
    stat1     = c( 1,   2,   3,   4 ),
    info1     = c("a", "b", "c", "d")
)
set_data <- IdVector(c("geneset1", "geneset2"))
mcols(set_data) <- DataFrame(
    stat1     = c( 100,   200 ),
    info1     = c("abc", "def")
)
base_sets <- BaseSets(relations_table, gene_data, set_data)
base_sets
```

The `elementData` and `setData` slots store metadata for each individual element and set that appears in the `relations` slot.
Those metadata can be accessed directly using the getter methods of the same name.

```{r}
elementData(base_sets)
setData(base_sets)
```

Note that relations between elements and sets are internally stored as an `r Biocpkg("S4Vectors")` `Hits` object.
This container efficiently represents a set of hits between a set of _left nodes_ and a set of _right nodes_, with optional metadata accompanying each hit.

```{r}
base_sets@relations
```

Conveniently, the `relations` accessor returns relations and associated metadata as a `DataFrame` substituting hits for their corresponding element and set identifiers.

```{r}
relations(base_sets)
```

## The _FuzzySets_ class

Classes derived from `Hits` may add additional constraints on the relations to define special types of relationships between elements and sets.

For instance, the `FuzzyHits` class is a direct extension of the `Hits` class where the metadata accompanying each relation must include at least a column called `"membership"` that holds the "membership function", a numeric value in the interval [0,1] that provides a measure of partial membership between elements and sets.

Simultaneously, the `FuzzySets` class is a direct extension of the `BaseSets` class where the `relations` slot must contain a `FuzzyHits`.
As such, `FuzzySets` can be constructed exactly like `BaseSets`, with the only additional constraint that the relations table contains a `"membership"` column with numeric values in the interval [0,1].

```{r, message=FALSE}
relations_table$membership <- round(runif(nrow(relations_table)), 2)
fuzzy_sets <- FuzzySets(relations_table, gene_data, set_data)
fuzzy_sets
```

The `membership` function associated with each relation can be directly accessed using the getter of the same name.

```{r}
membership(fuzzy_sets)
```

Conveniently, the `relations` accessor returns fuzzy relations as a `DataFrame` that includes the `"membership"` column, along with any other optional metadata accompanying the relations between elements and sets.

```{r}
relations(fuzzy_sets)
```

# Subsetting

The `subset` method can be applied to a `BaseSets` object, using a logical expression that may refer to the `"element"` and `"set"` columns as well as any metadata associated with the relations, indicating rows to keep.

```{r}
subset(base_sets, set == "geneset1" | element %in% c("C") | extra1 == "ABC")
```

Similarly, the `subset` method can be also applied to a `FuzzySets` object, where the logical expression may also refer to the additional `"membership"` metadata that is guaranted by the class validity method.

```{r}
subset(fuzzy_sets, set == "geneset2" & membership > 0.3)
```

# Converting to other formats

It is possible to extract the gene sets as a `list`, for use with functions such as `lapply`.

```{r}
as(fuzzy_sets, "list")
```

It is also possible to visualize membership between gene and gene sets as a matrix.

Notably, `BaseSets` objects produce a `logical` matrix of binary membership that indicates whether each element is associated at least once with each set:

```{r}
base_matrix <- as(base_sets, "matrix")
base_matrix
```

In contrast, `FuzzySets` objects produce a `double` matrix displaying the membership function for each relation.
Relations that are not described in the `FuzzySets` are filled with `NA`, to contrast with relations explictly associated with a membership function of 0.

```{r}
membership(fuzzy_sets)[1] <- 0
fuzzy_matrix <- as(fuzzy_sets, "matrix")
fuzzy_matrix
```

# Converting from other formats

## Matrix {#fromMatrix}

It is possible to convert incidence matrices into objects derived from the `BaseSets` class.

Notably, the `BaseSets` class is suitable for `logical` matrices indicating binary membership.

```{r}
as(base_matrix, "BaseSets")
```

Similarly, the `FuzzySets` class is suitable for `double` matrices indicating the membership function for each relation.
Importantly, relations described as `NA` are not imported into the `FuzzySets` object.
In contrast, relations with a membership function of 0 are imported and described as such.

```{r}
fuzzy_matrix[1, 1] <- 0
as(fuzzy_matrix, "FuzzySets")
```

# Additional information

## Count of relations, elements, and sets

The count of relations between elements and sets can be obtained using the `length` method.

```{r}
length(base_sets)
```

The count of unique elements and sets can be obtained using the `nElements` and `nSets` methods.

```{r}
nElements(base_sets)
nSets(base_sets)
```

The size of each gene set can be obtained using the `setLengths` method.

```{r}
setLengths(fuzzy_sets)
```

Conversely, the number of sets associated with each gene is returned by the `elementLengths` function.

```{r}
elementLengths(fuzzy_sets)
```

## Names of elements and sets

The identifiers of elements and sets can be inspected and renamed using `elementIds` and `setIds` methods.

```{r}
elementIds(base_sets) <- paste0("Gene", seq_len(nElements(base_sets)))
elementIds(base_sets)
setIds(base_sets) <- paste0("Geneset", seq_len(nSets(base_sets)))
setIds(base_sets)
```

# Importing and exporting sets

## Gene Matrix Transpose (GMT) Format

A common representation of gene sets is the GMT format, which is a non-rectangular format where each line is a set.
The first column is the name of the set, the second column is a description of the source of the set (such as a URL), and the third column onwards are the elements of the set, such that each set may have a variable number of elements.

Importing from and exporting to GMT files is performed using the generic `import` and `export` methods, which recognize the ".gmt" file extenson as a trigger to import from and export to the GMT file format.
Alternatively, the `import.gmt` and `import.gmt` functions may be used to explicitly export to the GMT file format.

Any object that inherits from the `BaseSets` class may be exported to the GMT file format.
However, any information that is not supported by the GMT file format will be lost during the export.
Reciprocally, the `import` function produces a `BaseSets` object, which adequately represents all the information present in the GMT file format.

```{r}
gmt_file <- system.file(package="unisets", "extdata", "example.gmt")
base_sets_from_gmt <- import(gmt_file)
base_sets_from_gmt
```

The additional metadata corresponding to the source (second column of the GMT) per set is also added as metadata corresponding to the sets, accessible via `setData`, which returns an `IdVector` class object.

```{r}
setData(base_sets_from_gmt)
```

To access the internal DataFrame representation, the accessor `mcols` can additionally be applied.

```{r}
mcols(setData(base_sets_from_gmt))
## elementMetadata(setData(base_sets_from_gmt)) # equivalent to above
```

To export `BaseSets` objects in GMT file format, the `export` generic may be used if the file extension is ".gmt".
Alternatively, data in GMT format may be exported to files with different extensions (e.g., ".txt") using the `export.gmt` function.
Note that if `"source"` heading is not found in the set metadata (i.e., `mcols(setData(x))`), this value will be filled with `"unisets"` in the exported file.

```{r}
tmp_file <- tempfile(fileext=".gmt")
export(base_sets_from_gmt, tmp_file)
```

# Additional information

Bug reports can be posted as issues in the `r Githubpkg("kevinrue/unisets")` GitHub repository.
The GitHub repository is the primary source for development versions of the package, where new functionality is added over time.
The authors appreciate well-considered suggestions for improvements or new features, or even better, pull requests.

If you use `r Githubpkg("kevinrue/unisets")` for your analysis, please cite it as shown below:

```{r citation}
citation("unisets")
```

# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
